---
title: "analysis 5 sycophancy"
output: word_document
date: "2025-12-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,      # Don't show code
  warning = FALSE,   # Don't show warnings
  message = FALSE   # Don't show messages

)

library(tidyverse)
library(jsonlite)
library(purrr)
library(stringr)
library(lmerTest)
library(papaja)
library(broom)
library(forcats)
library(patchwork)
library(fs)
library(ggplot2)
library(matrixStats)

# Create output directory
fs::dir_create("figures/sycophancy", recurse = TRUE)

```


```{r load data}
source('helpers.R')
source('loadAndFormatSycophancyData.R')
```

```{r reshape}

# Reshape for scatterplots
# Note: Round removed_response to avoid floating point precision issues in pivot_wider
Qwen.sycophancy_wide <- Qwen.sycophancy_processed %>%
  select(scenario_id, my_version, prompt, response, removed_response, marginal_response) %>%
  mutate(removed_response = round(removed_response, 14)) %>%
  pivot_wider(
    names_from = my_version,
    values_from = c(response, marginal_response),
    id_cols = c(scenario_id, prompt, removed_response)
  ) %>%
  rename(
    response_when_A_is_me = response_A,
    response_when_B_is_me = response_B,
    marginal_when_A_is_me = marginal_response_A,
    marginal_when_B_is_me = marginal_response_B
  ) %>%
  mutate(model = "Qwen")

GPT.sycophancy_wide <- GPT.sycophancy_processed %>%
  select(scenario_id, my_version, prompt, response, removed_response, marginal_response) %>%
  mutate(removed_response = round(removed_response, 14)) %>%
  pivot_wider(
    names_from = my_version,
    values_from = c(response, marginal_response),
    id_cols = c(scenario_id, prompt, removed_response)
  ) %>%
  rename(
    response_when_A_is_me = response_A,
    response_when_B_is_me = response_B,
    marginal_when_A_is_me = marginal_response_A,
    marginal_when_B_is_me = marginal_response_B
  ) %>%
  mutate(model = "GPT")

# Combine
combined_sycophancy_wide <- bind_rows(Qwen.sycophancy_wide, GPT.sycophancy_wide)

# Check dimensions
cat(sprintf("Qwen wide: %d rows (should be 60 scenarios × 5 prompts = 300)\n", 
            nrow(Qwen.sycophancy_wide)))
cat(sprintf("GPT wide: %d rows (should be 60 scenarios × 5 prompts = 300)\n", 
            nrow(GPT.sycophancy_wide)))

# Preview
head(combined_sycophancy_wide)

# Test scatterplot for Default prompt
GPT.sycophancy_wide %>%
  filter(prompt == "Default") %>%
  ggplot(aes(x = response_when_B_is_me, y = response_when_A_is_me)) +
  geom_hline(yintercept = 0, linetype = "solid", color = "gray30") +
  geom_vline(xintercept = 0, linetype = "solid", color = "gray30") +
  geom_point(color = "gray50", alpha = 0.6, size = 2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  labs(
    title = "GPT-4.1: Default (Sycophancy)",
    x = "Response when A is 'them' (user is B)",
    y = "Response when A is 'me' (user is A)"
  ) +
  theme_minimal() +
  theme(aspect.ratio = 1)

```

```{r plot}

create_sycophancy_plot_with_coef <- function(data, model_name, prompt_name, prompt_color) {
  
  # Calculate mean displacements for this prompt and Default
  summary_stats <- data %>%
    filter(model == model_name, prompt %in% c(prompt_name, "Default")) %>%
    group_by(prompt) %>%
    summarise(
      mean_displacement = mean(response_when_A_is_me - response_when_B_is_me),
      se_displacement = sd(response_when_A_is_me - response_when_B_is_me) / sqrt(n()),
      t_displacement = mean_displacement / se_displacement,
      p_displacement = 2 * pt(abs(t_displacement), n() - 1, lower.tail = FALSE),
      mean_marginal_displacement = mean(marginal_when_A_is_me - marginal_when_B_is_me),
      se_marginal_displacement = sd(marginal_when_A_is_me - marginal_when_B_is_me) / sqrt(n()),
      t_marginal = mean_marginal_displacement / se_marginal_displacement,
      p_marginal = 2 * pt(abs(t_marginal), n() - 1, lower.tail = FALSE),
      .groups = "drop"
    ) %>%
    mutate(
      stars_displacement = case_when(
        p_displacement < 0.001 ~ "***",
        p_displacement < 0.01 ~ "**",
        p_displacement < 0.05 ~ "*",
        TRUE ~ ""
      ),
      stars_marginal = case_when(
        p_marginal < 0.001 ~ "***",
        p_marginal < 0.01 ~ "**",
        p_marginal < 0.05 ~ "*",
        TRUE ~ ""
      )
    )
  
  default_stats <- summary_stats %>% filter(prompt == "Default")
  prompt_stats <- summary_stats %>% filter(prompt == prompt_name)
  
  # Create scatterplot with title
  scatter <- data %>%
    filter(model == model_name, prompt == prompt_name) %>%
    ggplot(aes(x = response_when_B_is_me, y = response_when_A_is_me)) +
    geom_hline(yintercept = 0, linetype = "solid", color = "gray30") +
    geom_vline(xintercept = 0, linetype = "solid", color = "gray30") +
    geom_abline(slope = 1, intercept = 0, color = "black") +
    # Prompt points in color
    geom_point(color = prompt_color, alpha = 0.5, size = 2) +
    # Marginal points in black
    geom_point(aes(x = marginal_when_B_is_me, y = marginal_when_A_is_me),
               color = "black", alpha = 0.7, size = 2) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "white") +
    labs(
      title = prompt_name,
      x = "Party A is 'them'",
      y = "Party A is 'me'"
    ) +
    theme_minimal() +
    theme(
      aspect.ratio = 1,
      plot.title = element_text(hjust = 0.5, size = 12, face = "bold")
    )
  
displacement_data <- bind_rows(
  default_stats %>% mutate(condition = "Default", color = "gray50"),
  prompt_stats %>% mutate(condition = prompt_name, color = prompt_color),
  prompt_stats %>% 
    mutate(condition = paste0(prompt_name, " (blinded)"), color = "black",
           mean_displacement = mean_marginal_displacement,
           se_displacement = se_marginal_displacement,
           stars_displacement = stars_marginal)
) %>%
  mutate(
    condition = factor(condition, levels = c("Default", prompt_name, paste0(prompt_name, " (blinded)"))),
    star_y = mean_displacement + se_displacement + 0.5 * max(abs(mean_displacement))
  )
  
# Create displacement plot
displacement_plot <- displacement_data %>%
  ggplot(aes(x = condition, y = mean_displacement)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  # Original in color
  geom_point(data = displacement_data %>% filter(condition != paste0(prompt_name, " (blinded)")),
             aes(color = color), size = 3, alpha = 0.5) +
  geom_errorbar(data = displacement_data %>% filter(condition != paste0(prompt_name, " (blinded)")),
                aes(ymin = mean_displacement - se_displacement,
                    ymax = mean_displacement + se_displacement,
                    color = color),
                width = 0.2, linewidth = 0.6, alpha = 0.5) +
  # Marginal in black
  geom_point(data = displacement_data %>% filter(condition == paste0(prompt_name, " (blinded)")),
             aes(color = color), size = 3) +
  geom_errorbar(data = displacement_data %>% filter(condition == paste0(prompt_name, " (blinded)")),
                aes(ymin = mean_displacement - se_displacement,
                    ymax = mean_displacement + se_displacement,
                    color = color),
                width = 0.2, linewidth = 0.6) +
  # Stars
  geom_text(data = displacement_data,
            aes(x = condition, y = star_y, label = stars_displacement, color = color),
            size = 4, fontface = "bold", vjust = 0, inherit.aes = FALSE) +
  scale_color_identity() +
  labs(x = NULL, y = "Sycophancy") +
  theme_minimal() +
  coord_cartesian(clip = "off") +
  theme(
    aspect.ratio = 0.3,
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.y = element_text(size = 8),
    panel.grid.major.x = element_blank(),
    plot.margin = margin(t = 15, r = 1, b = 2, l = 1)
  )
  
  # Combine
  combined <- (scatter / displacement_plot) + plot_layout(heights = c(3, 1))
  
  return(combined)
}

# Define colors
prompt_colors <- c(
  "Ignore" = "#4DAF4A",
  "Don't Be Sycophantic" = "#377EB8",
  "If You Didn't Know" = "#984EA3",
  "Neutral" = "#FF7F00"
)

# Create plots for GPT
gpt_ignore <- create_sycophancy_plot_with_coef(combined_sycophancy_wide, "GPT", "Ignore", prompt_colors["Ignore"])
gpt_dont <- create_sycophancy_plot_with_coef(combined_sycophancy_wide, "GPT", "Don't Be Sycophantic", prompt_colors["Don't Be Sycophantic"])
gpt_if <- create_sycophancy_plot_with_coef(combined_sycophancy_wide, "GPT", "If You Didn't Know", prompt_colors["If You Didn't Know"])

# Create plots for Qwen
qwen_ignore <- create_sycophancy_plot_with_coef(combined_sycophancy_wide, "Qwen", "Ignore", prompt_colors["Ignore"])
qwen_dont <- create_sycophancy_plot_with_coef(combined_sycophancy_wide, "Qwen", "Don't Be Sycophantic", prompt_colors["Don't Be Sycophantic"])
qwen_if <- create_sycophancy_plot_with_coef(combined_sycophancy_wide, "Qwen", "If You Didn't Know", prompt_colors["If You Didn't Know"])

# Combine separately for each model
library(patchwork)
gpt_grid <- gpt_ignore | gpt_dont | gpt_if
qwen_grid <- qwen_ignore | qwen_dont | qwen_if

# Save using models from helpers.R
ggsave(sprintf('figures/sycophancy/%s_sycophancy.png', models[["GPT"]]), plot = gpt_grid, height = 5, width = 15, dpi = 300)
ggsave(sprintf('figures/sycophancy/%s_sycophancy.png', models[["Qwen"]]), plot = qwen_grid, height = 5, width = 15, dpi = 300)

```

```{r bar plots}

# Calculate mean absolute displacement for all model × prompt × blinded combinations
sycophancy_mean_abs_displacement <- combined_sycophancy_wide %>%
  pivot_longer(
    cols = c(response_when_A_is_me, response_when_B_is_me, 
             marginal_when_A_is_me, marginal_when_B_is_me),
    names_to = "variable",
    values_to = "value"
  ) %>%
  mutate(
    response_type = ifelse(grepl("marginal", variable), "blinded", "original"),
    user_side = ifelse(grepl("A_is_me", variable), "A", "B")
  ) %>%
  select(model, prompt, scenario_id, response_type, user_side, value, removed_response) %>%
  pivot_wider(
    names_from = user_side,
    values_from = value
  ) %>%
  mutate(
    displacement = A - B,
    abs_displacement = abs(displacement)
  ) %>%
  group_by(model, prompt, response_type) %>%
  summarise(
    mean_abs_displacement = mean(abs_displacement),
    se_abs_displacement = sd(abs_displacement) / sqrt(n()),
    .groups = "drop"
  )

# View the results
print(sycophancy_mean_abs_displacement)

# Check dimensions (should be 2 models × 5 prompts × 2 response types = 20 rows)
cat(sprintf("Total rows: %d\n", nrow(sycophancy_mean_abs_displacement)))

# Prepare data for barplot
barplot_data_original <- sycophancy_mean_abs_displacement %>%
  filter(response_type == "original")

barplot_data_blinded <- sycophancy_mean_abs_displacement %>%
  filter(response_type == "blinded")

# Define colors
prompt_colors_bar <- c(
  "Default" = "gray75",
  "Ignore" = "#4DAF4A",
  "Don't Be Sycophantic" = "#377EB8",
  "If You Didn't Know" = "#984EA3",
  "Neutral" = "#FF7F00"
)

# Create barplot function
create_sycophancy_barplot <- function(data_original, data_blinded, model_name, output_path) {
  
  # Filter for this model
  original <- data_original %>% 
    filter(model == model_name) %>%
    mutate(
      fill_color = prompt_colors_bar[prompt],
      prompt = factor(prompt, levels = rev(c("Default", "Ignore", "Don't Be Sycophantic", 
                                              "If You Didn't Know", "Neutral"))),
      label_text = case_when(
        prompt == "Default" ~ "Default",
        prompt == "Ignore" ~ "Ignore",
        prompt == "Don't Be Sycophantic" ~ "Don't Be Sycophantic",
        prompt == "If You Didn't Know" ~ "If You Didn't Know",
        prompt == "Neutral" ~ "Neutral"
      )
    )
  
  blinded <- data_blinded %>%
    filter(model == model_name) %>%
    mutate(prompt = factor(prompt, levels = rev(c("Default", "Ignore", "Don't Be Sycophantic", 
                                                   "If You Didn't Know", "Neutral"))))
  
  p <- ggplot(original, aes(x = prompt, y = mean_abs_displacement)) +
    # Original bars in color
    geom_bar(aes(fill = fill_color), stat = "identity", width = 0.7) +
    scale_fill_identity() +
    # Blinded bars overlaid in black
    geom_bar(data = blinded, stat = "identity", width = 0.7, fill = "black") +
    # Error bars for original
    geom_errorbar(aes(ymin = mean_abs_displacement - se_abs_displacement,
                      ymax = mean_abs_displacement + se_abs_displacement),
                  width = 0.2, linewidth = 0.8) +
    # Error bars for blinded
    geom_errorbar(data = blinded,
                  aes(ymin = mean_abs_displacement - se_abs_displacement,
                      ymax = mean_abs_displacement + se_abs_displacement),
                  width = 0.2, linewidth = 0.8) +
    # Text labels
    geom_text(aes(y = 0.5, label = label_text),
              hjust = 0, size = 3, fontface = "bold", color = "white") +
    scale_y_continuous(expand = c(0, 0),
                       limits = c(0, max(original$mean_abs_displacement) * 1.1)) +
    labs(x = NULL, y = "Mean absolute sycophancy") +
    coord_flip() +
    theme_minimal() +
    theme(
      legend.position = "none",
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      panel.grid.major.y = element_blank()
    )
  
  ggsave(output_path, plot = p, height = 5, width = 6, dpi = 300)
  
  return(p)
}

# Create barplots using models from helpers.R
gpt_barplot <- create_sycophancy_barplot(
  barplot_data_original %>% filter(prompt !='Neutral'),
  barplot_data_blinded %>% filter(prompt !='Neutral'),
  "GPT",
  sprintf('figures/sycophancy/%s_sycophancy_barplot.png', models[["GPT"]])
)

qwen_barplot <- create_sycophancy_barplot(
  barplot_data_original %>% filter(prompt !='Neutral'),
  barplot_data_blinded %>% filter(prompt !='Neutral'),
  "Qwen",
  sprintf('figures/sycophancy/%s_sycophancy_barplot.png', models[["Qwen"]])
)

print(gpt_barplot)
print(qwen_barplot)
```

```{r complete plots}

# Create complete figure for each model
create_sycophancy_complete_figure <- function(data, mean_abs_data_original, mean_abs_data_blinded, 
                                              model_name, output_path) {
  
  # Define colors
  prompt_colors <- c(
    "Ignore" = "#4DAF4A",
    "Don't Be Sycophantic" = "#377EB8",
    "If You Didn't Know" = "#984EA3"
  )
  
  prompt_colors_bar <- c(
    "Default" = "gray",
    "Ignore" = "#4DAF4A",
    "Don't Be Sycophantic" = "#377EB8",
    "If You Didn't Know" = "#984EA3"
  )
  
  # Create the three scatter+displacement plots
  p1 <- create_sycophancy_plot_with_coef(data, model_name, "Don't Be Sycophantic", 
                                         prompt_colors["Don't Be Sycophantic"])
  p2 <- create_sycophancy_plot_with_coef(data, model_name, "Ignore", 
                                         prompt_colors["Ignore"])
  p3 <- create_sycophancy_plot_with_coef(data, model_name, "If You Didn't Know", 
                                         prompt_colors["If You Didn't Know"])
  
  # Create barplot
  original <- mean_abs_data_original %>% 
    filter(model == model_name, prompt %in% c("Default", "Ignore", "Don't Be Sycophantic", "If You Didn't Know")) %>%
    mutate(
      fill_color = prompt_colors_bar[prompt],
      prompt = factor(prompt, levels = rev(c("Default", "Ignore", "Don't Be Sycophantic", "If You Didn't Know"))),
      label_text = case_when(
        prompt == "Default" ~ "Default",
        prompt == "Ignore" ~ "Ignore",
        prompt == "Don't Be Sycophantic" ~ "Don't Be Sycophantic",
        prompt == "If You Didn't Know" ~ "If You Didn't Know"
      )
    )
  
  blinded <- mean_abs_data_blinded %>%
    filter(model == model_name, prompt %in% c("Default", "Ignore", "Don't Be Sycophantic", "If You Didn't Know")) %>%
    mutate(prompt = factor(prompt, levels = rev(c("Default", "Ignore", "Don't Be Sycophantic", "If You Didn't Know"))))
  
  p4 <- ggplot(original, aes(x = prompt, y = mean_abs_displacement)) +
    # Original bars in color
    geom_bar(aes(fill = fill_color), stat = "identity", width = 0.7) +
    scale_fill_identity() +
    # Blinded bars overlaid in black
    geom_bar(data = blinded, stat = "identity", width = 0.7, fill = "black") +
    # Error bars for original
    geom_errorbar(aes(ymin = mean_abs_displacement - se_abs_displacement,
                      ymax = mean_abs_displacement + se_abs_displacement),
                  width = 0.2, linewidth = 0.8) +
    # Error bars for blinded
    geom_errorbar(data = blinded,
                  aes(ymin = mean_abs_displacement - se_abs_displacement,
                      ymax = mean_abs_displacement + se_abs_displacement),
                  width = 0.2, linewidth = 0.8) +
    # Text labels
    geom_text(aes(y = 0.5, label = label_text),
              hjust = 0, size = 3, fontface = "bold", color = "white") +
    scale_y_continuous(expand = c(0, 0),
                       limits = c(0, max(original$mean_abs_displacement) * 1.1)) +
    labs(x = NULL, y = "Mean absolute sycophancy") +
    coord_flip() +
    theme_minimal() +
    theme(
      legend.position = "none",
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      panel.grid.major.y = element_blank()
    )
  
  # Combine in 2×2 grid
  combined <- (p1 | p2) / (p3 | p4)
  
  # Save
  ggsave(output_path, plot = combined, height = 7, width = 6, dpi = 300)
  
  return(combined)
}

# Create complete figures using models from helpers.R
gpt_complete <- create_sycophancy_complete_figure(
  combined_sycophancy_wide,
  barplot_data_original,
  barplot_data_blinded,
  "GPT",
  sprintf('figures/sycophancy/%s_sycophancy_complete.png', models[["GPT"]])
)

qwen_complete <- create_sycophancy_complete_figure(
  combined_sycophancy_wide,
  barplot_data_original,
  barplot_data_blinded,
  "Qwen",
  sprintf('figures/sycophancy/%s_sycophancy_complete.png', models[["Qwen"]])
)

```

```{r deferral}
deferral_by_side <- combined_sycophancy %>%
  mutate(
    deferral_when_favourable = ifelse(
      my_version=='A', pA_given_A, 1-pA_given_B
      ),
    deferral_when_unfavourable = ifelse(
      my_version=='A',1-pA_given_B, pA_given_A
    ),
  )

# Approach 1: Average within scenarios first (df = 59)
deferral_ttests_scenario <- deferral_by_side %>%
  filter(prompt != "Default") %>%
  group_by(model, prompt, scenario_id) %>%
  summarise(
    deferral_when_favourable = mean(deferral_when_favourable),
    deferral_when_unfavourable = mean(deferral_when_unfavourable),
    .groups = "drop"
  ) %>%
  group_by(model, prompt) %>%
  summarise(
    mean_favorable = mean(deferral_when_favourable),
    mean_unfavorable = mean(deferral_when_unfavourable),
    t_statistic = t.test(deferral_when_favourable, deferral_when_unfavourable, 
                         paired = TRUE)$statistic,
    p_value = t.test(deferral_when_favourable, deferral_when_unfavourable, 
                     paired = TRUE)$p.value,
    df = t.test(deferral_when_favourable, deferral_when_unfavourable, 
                paired = TRUE)$parameter,
    .groups = "drop"
  ) %>%
  mutate(
    significance = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",
      TRUE ~ ""
    )
  )

```

Both models mostly deferrred to the blinded, counterfactual selves (Qwen did so on 
`r c(combined_sycophancy%>%filter(model=='Qwen')%>%pull(pA_given_A),1-combined_sycophancy%>%filter(model=='GPT')%>%pull(pA_given_B))%>%mean()%>%printnum()` of all cases, GPT on `r c(combined_sycophancy%>%filter(model=='GPT')%>%pull(pA_given_A),1-combined_sycophancy%>%filter(model=='GPT')%>%pull(pA_given_B))%>%mean()%>%printnum()`). As a result, their responses were more aligned with that of a truly blinded model (Fig., black bars). And yet, the two models deferred to their blinded selves significantly more when its response was favourable to the user (Qwen: `r deferral_by_side%>%filter(model=='Qwen')%>%pull(deferral_when_favourable)%>%mean()%>%printnum()`, GPT: `r deferral_by_side%>%filter(model=='GPT')%>%pull(deferral_when_favourable)%>%mean()%>%printnum()`), compared to not (Qwen: `r deferral_by_side%>%filter(model=='Qwen')%>%pull(deferral_when_unfavourable)%>%mean()%>%printnum()`, GPT: `r deferral_by_side%>%filter(model=='GPT')%>%pull(deferral_when_unfavourable)%>%mean()%>%printnum()`). This difference in deferral rates was significant for all cases apart from Qwen, when instructed to ignore the user's identity, and GPT, when  instructed not ot be sycophantic. It resulted in significant sycophancy for all interventions ($p<0.01$ for all), even if much smaller than with any other intervention (see Fig. , black markers). This selective deferral to the blinded model is revealing: we take it to mean that models are sometimes knowingly sycophantic, giving user-serving responses even when instructed to avoid doing so and with direct access to a counterfactual unbiased response.