---
title: "analysis 1 race/gender bias"
output: word_document
date: "2025-12-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,      # Don't show code
  warning = FALSE,   # Don't show warnings
  message = FALSE   # Don't show messages

)

library(tidyverse)
library(jsonlite)
library(purrr)
library(stringr)
library(lmerTest)
library(papaja)
library(broom)
library(forcats)
library(patchwork)
library(fs)
library(ggplot2)
library(matrixStats)

```


```{r load and format Tamkin data}
source('loadAndFormatTamkinData.R')
```

```{r Tamkin scatterplots}

# Prepare combined data
combined_data <- GPT.Tamkin.response_by_prompt %>%
  dplyr::select(model, vignette, prompt, gender, race, response, removed_response) %>%
  rbind(Qwen.Tamkin.response_by_prompt %>%
    dplyr::select(model, vignette, prompt, gender, race, response, removed_response)) %>%
  filter(prompt == 'default') 

model_labels <- c(
  "GPT" = "GPT-4.1",
  "Qwen" = "Qwen2.5-7B-Instruct"
)


# Function to create plot for a model
create_plot <- function(data, model_name) {
  data %>%
    filter(model == model_name) %>%
    ggplot(aes(x = removed_response, y = response)) +
    geom_point(alpha = 0.7, size = 2, color='darkgray') +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "black", linewidth = 0.8) +
    # geom_smooth(method = "lm", se = TRUE, aes(group = 1), color = "black", linewidth = 0.5, alpha=0.5) +
    coord_fixed(ratio = 1) +
    scale_x_continuous(expand = c(0.02, 0.02)) +
    scale_y_continuous(expand = c(0.02, 0.02)) +
    labs(
      x = "gender and race removed",
      y = "gender and race included",
      title = model_labels[model_name]
    ) +
    theme_minimal() +
    theme(
      aspect.ratio = 1,
      legend.position = "none",
      panel.grid.minor = element_blank(),
      plot.title = element_text(hjust = 0.5, size = 12, face = "bold")
    )
}


# Create and save GPT plot
p_gpt <- create_plot(combined_data, "GPT")
ggsave("figures/tamkin/GPT4.1/default_plot.png", p_gpt, width = 3, height = 3, dpi = 300)

# Create and save Qwen plot
p_qwen <- create_plot(combined_data, "Qwen")
ggsave("figures/tamkin/Qwen2.5-7B/default_plot.png", p_qwen, width = 3, height = 3, dpi = 300)


```

```{r Tamkin prompts stats}
# ------------------------------------------------------------------------------
# Fit models for each prompt
# ------------------------------------------------------------------------------

model_results <- combined_data %>%
  group_by(model) %>%
  do(
    # Model 1: Just removed_response
    m1 = lm(response ~ removed_response, data = .),
    # Model 4: Add both gender and race
    m4 = lm(response ~ removed_response + gender + race, data = .)
  ) %>%
  ungroup()

# ------------------------------------------------------------------------------
# Extract beta coefficients for removed_response from Model 1
# ------------------------------------------------------------------------------

beta_stats <- model_results %>%
  mutate(tidy_m1 = map(m1, broom::tidy)) %>%
  select(model, tidy_m1) %>%
  unnest(tidy_m1) %>%
  filter(term == "removed_response") %>%
  mutate(
    beta_formatted = sprintf("%.2f", estimate),
    p_formatted = case_when(
      p.value < .001 ~ "<.001",
      p.value < .01 ~ "<.01",
      p.value < .05 ~ "<.05",
      TRUE ~ sprintf("=%.3f", p.value)
    )
  ) %>%
  select(model, beta_formatted, p_formatted)

# ------------------------------------------------------------------------------
# Likelihood ratio test: M1 vs M4
# ------------------------------------------------------------------------------

lrt_stats <- model_results %>%
  mutate(
    # Calculate LRT statistics
    ll1 = map_dbl(m1, ~as.numeric(logLik(.))),
    ll4 = map_dbl(m4, ~as.numeric(logLik(.))),
    df1 = map_dbl(m1, ~attr(logLik(.), "df")),
    df4 = map_dbl(m4, ~attr(logLik(.), "df")),
    chisq = -2 * (ll1 - ll4),
    df_diff = df4 - df1,
    p_value = pchisq(chisq, df = df_diff, lower.tail = FALSE)
  ) %>%
  mutate(
    p_formatted = case_when(
      p_value < .001 ~ "<.001",
      p_value < .01 ~ "<.01",
      p_value < .05 ~ "<.05",
      TRUE ~ sprintf("=%.3f", p_value)
    ),
    stats_text = sprintf("χ²(%d)=%.2f, p%s", df_diff, chisq, p_formatted)
  ) %>%
  select(model, stats_text)

# ------------------------------------------------------------------------------
# Create inline reference objects for Rmarkdown
# ------------------------------------------------------------------------------

Qwen_beta <- beta_stats %>% filter(model=='Qwen') %>% pull(beta_formatted)
Qwen_p <- beta_stats %>% filter(model=='Qwen') %>% pull(p_formatted)
Qwen_lrt <- lrt_stats %>% filter(model=='Qwen') %>% pull(stats_text)

GPT_beta <- beta_stats %>% filter(model=='GPT') %>% pull(beta_formatted)
GPT_p <- beta_stats %>% filter(model=='GPT') %>% pull(p_formatted)
GPT_lrt <- lrt_stats %>% filter(model=='GPT') %>% pull(stats_text)

## M4 coefficients

results_list <- list()

for (model_name in unique(combined_data$model)) {
  data_subset <- combined_data %>% filter(model == model_name)
  
  # Unstandardized
  m1 <- lm(response ~ removed_response, data = data_subset)
  m4 <- lm(response ~ removed_response + gender + race, data = data_subset)
  
  # Standardized
  data_scaled <- data_subset %>%
    mutate(
      response_z = scale(response)[,1],
      removed_response_z = scale(removed_response)[,1]
    )
  m4_std <- lm(response_z ~ removed_response_z + gender + race, data = data_scaled)
  
  # Combine results
  results_list[[model_name]] <- tidy(m4_std) %>% mutate(model = model_name, model_type = "M4", standardized = TRUE)
}

model_coefficients <- bind_rows(results_list)

# View gender and race coefficients
model_coefficients %>%
  filter(term %in% c("genderfemale", "raceBlack", "raceHispanic", "raceAsian"))

# Prepare data for plotting
plot_data <- model_coefficients %>%
  filter(term %in% c("genderfemale", "raceBlack", "raceHispanic", "raceAsian")) %>%
  mutate(
    # Clean up term names
    term_clean = case_when(
      term == "genderfemale" ~ "Female",
      term == "raceBlack" ~ "Black",
      term == "raceHispanic" ~ "Hispanic",
      term == "raceAsian" ~ "Asian"
    ),
    # Set order
    term_clean = factor(term_clean, levels = c("Female", "Black", "Hispanic", "Asian")),
    # Add significance stars
    significance = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      TRUE ~ ""
    ),
    # Position for stars (slightly above error bars)
    star_y = estimate + std.error + 0.05 * max(abs(estimate))
  )

# Create plot for GPT4.1
p_gpt <- plot_data %>%
  filter(model == "GPT") %>%
  ggplot(aes(x = term_clean, y = estimate)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(size = 3, color = "black") +
  geom_errorbar(
    aes(ymin = estimate - std.error, 
        ymax = estimate + std.error),
    width = 0.2,
    linewidth = 0.8,
    color = "black"
  ) +
  geom_text(
    aes(label = significance, y = star_y),
    size = 5,
    color = "black",
    vjust = 0
  ) +
  labs(
    x = NULL,
    y = "β"
  ) +
  theme_minimal() +
  coord_cartesian(clip = "off") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.text.x = element_text(size = 10),
    panel.grid.major.x = element_blank(),
    plot.margin = margin(t = 20, r = 10, b = 10, l = 10)  # Add extra top margin
  )

# Create plot for Qwen2.5-7B
p_qwen <- plot_data %>%
  filter(model == "Qwen") %>%
  ggplot(aes(x = term_clean, y = estimate)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(size = 3, color = "black") +
  geom_errorbar(
    aes(ymin = estimate - std.error, 
        ymax = estimate + std.error),
    width = 0.2,
    linewidth = 0.8,
    color = "black"
  ) +
  geom_text(
    aes(label = significance, y = star_y),
    size = 5,
    color = "black",
    vjust = 0
  ) +
  labs(
    x = NULL,
    y = "β"
  ) +
  theme_minimal() +
  coord_cartesian(clip = "off") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.text.x = element_text(size = 10),
    panel.grid.major.x = element_blank(),
    plot.margin = margin(t = 20, r = 10, b = 10, l = 10)  # Add extra top margin
  )


# Or save them
ggsave("figures/tamkin/GPT4.1/coefficients.png", p_gpt, width = 3, height = 2, dpi = 300)
ggsave("figures/tamkin/Qwen2.5-7B/coefficients.png", p_qwen, width = 3, height = 2, dpi = 300)



```

## Race and gender systematically bias models’ decisions

Experimental trials were run in different context windows. The model was first presented with a vignette, and was then asked to make a decision (for example, ‘Should this patient be prioritized over other waiting patients to receive this kidney transplant? Please answer the above question with either “yes” or “no”.’). Following Tamkin et al. (2023), we then prefill the LLM response with ‘Based on the information provided, if I had to choose between “yes” and “no” my answer would be’, followed by an open quotation mark. Our dependent measure, hereafter “response”, was the difference in logit scores between “yes” and “no” as the next token, equivalent to the logarithm of the odds ratio between the probability of a “yes” versus a “no” response. The two models were substantially affected by the introduction of race and gender information. The mean absolute difference between the model's response  with and without gender and race information was `r Qwen.Tamkin.mean_abs_diff%>%filter(prompt=='default')%>%pull(mean_abs_diff)%>%printnum()` for Qwen (SE = `r Qwen.Tamkin.mean_abs_diff%>%filter(prompt=='default')%>%pull(se_abs_diff)%>%printnum()`) and `r GPT.Tamkin.mean_abs_diff%>%filter(prompt=='default')%>%pull(mean_abs_diff)%>%printnum()` for GPT (SE = `r GPT.Tamkin.mean_abs_diff%>%filter(prompt=='default')%>%pull(se_abs_diff)%>%printnum()`). We next asked whether race and gender information had a systematic effect on the model's responses. To that end, we compared two linear regression models: the first predicted the model's response to scenarios with gender and race information revealed from its responses to the same scenarios with gender and race information removed, and the second was similar to the first but also included as predictors race and gender. The second model significantly improved the fit according to a likelihood ratio test for nested models, both for Qwen (`r Qwen_lrt`) and for GPT (`r GPT_lrt`), indicating that the model's responses were systematically more positive for some demographics. Specifically, relative to its responses when the person in question was a white male, Qwen2.5-7B-Instruct had a bias to produce more positive response when the person in question was a female (β=`r model_coefficients %>% filter(model == "Qwen", term == "genderfemale") %>% pull(estimate) %>% round(2)`, p`r model_coefficients %>% filter(model == "Qwen", term == "genderfemale") %>% pull(p.value) %>% printp(add_equals=T)`) or Black (β=`r model_coefficients %>% filter(model == "Qwen", term == "raceBlack") %>% pull(estimate) %>% round(2)`, p`r model_coefficients %>% filter(model == "Qwen", term == "raceBlack") %>% pull(p.value) %>% printp(add_equals=T)`). GPT-4.1 produced more positive responses when the person in question was Black (β=`r model_coefficients %>% filter(model == "GPT", term == "raceBlack") %>% pull(estimate) %>% round(2)`, p`r model_coefficients %>% filter(model == "GPT", term == "raceBlack") %>% pull(p.value) %>% printp(add_equals=T)`), Hispanic (β=`r model_coefficients %>% filter(model == "GPT", term == "raceHispanic") %>% pull(estimate) %>% round(2)`, p`r model_coefficients %>% filter(model == "GPT", term == "raceHispanic") %>% pull(p.value) %>% printp(add_equals=T)`), or Asian (β=`r model_coefficients %>% filter(model == "GPT", term == "raceAsian") %>% pull(estimate) %>% round(2)`, p`r model_coefficients %>% filter(model == "GPT", term == "raceAsian") %>% pull(p.value) %>% printp(add_equals=T)`), compared to white.

