---
title: "analysis 3 tool use"
output: word_document
date: "2025-12-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,      # Don't show code
  warning = FALSE,   # Don't show warnings
  message = FALSE   # Don't show messages

)

library(tidyverse)
library(jsonlite)
library(purrr)
library(stringr)
library(lmerTest)
library(papaja)
library(broom)
library(forcats)
library(patchwork)
library(fs)
library(ggplot2)
library(matrixStats)

```


```{r load data}
source('helpers.R')
source('parse_tool_generations.R')

# Load bias data
Qwen.Bias.YN.df <- load_bias_data("qwen")
GPT.Bias.YN.df <- load_bias_data("gpt")

# Load tool use data (CSV + generation parsing for Qwen)
tool_data <- load_tool_use_data("both")
Qwen.Bias.tool_use_combined <- tool_data$qwen
GPT.Bias.tool_use_combined <- tool_data$gpt
```


```{r plot}
# Combine the two dataframes
# Note: vignette column already exists from helpers (renamed from vignette)
all_tool_calls <- bind_rows(
  GPT.Bias.tool_use_combined,
  Qwen.Bias.tool_use_combined
)

# Define labels and colors
prompt_labels <- c(
  default = "Default",
  dont_discriminate = "Don't Discriminate",
  ignore = "Ignore",
  if_you_didnt_know = "If You Didn't Know",
  removed = "Removed"
)

prompt_colors <- c(
  default = "gray",           # Red
  dont_discriminate = "#377EB8", # Blue
  ignore = "#4DAF4A",            # Green
  if_you_didnt_know = "#984EA3", # Purple
  removed = "black"               # Gray
)

# Calculate tool use counts for GPT
gpt_counts_df <- all_tool_calls %>%
  filter(model == 'GPT') %>%
  group_by(prompt) %>%
  summarise(prop_tool_use = mean(tool_call))
         
# Calculate tool use counts for Qwen
qwen_counts_df <- all_tool_calls %>%
  filter(model == 'Qwen') %>%
  group_by(prompt) %>%
  summarise(prop_tool_use = mean(tool_prob))

# Create GPT tool use plot
ggplot(gpt_counts_df %>%
         filter(prompt %in% c('default','dont_discriminate','if_you_didnt_know','ignore','removed')), aes(x = prompt, y = prop_tool_use, fill = prompt)) +
  geom_col() +
  scale_fill_manual(values = prompt_colors, labels = prompt_labels) +
  scale_x_discrete(labels = prompt_labels) +
  labs(
    x = "",
    y = "p(self-call)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )

ggsave('figures/self-calls/GPT4.1/proportion_tool_use_demographic_bias.png', height = 2.5, width = 4, dpi = 300)

# Create Qwen tool use plot
ggplot(qwen_counts_df %>%
         filter(prompt %in% c('default','dont_discriminate','if_you_didnt_know','ignore','removed')), aes(x = prompt, y = prop_tool_use, fill = prompt)) +
  geom_col() +
  scale_fill_manual(values = prompt_colors, labels = prompt_labels) +
  scale_x_discrete(labels = prompt_labels) +
  labs(
    x = "",
    y = "p(self-call)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )

ggsave('figures/self-calls/Qwen2.5-7B/proportion_tool_use_demographic_bias.png', height = 2.5, width = 4, dpi = 300)

# Calculate proportions of demographic information inclusion
inclusion_props <- all_tool_calls %>%
  filter(!is.na(tool_call_text)) %>%  # Only look at rows with tool calls
  group_by(prompt, model) %>%
  summarise(
    n = n(),
    prop_includes_race = mean(includes_race, na.rm = TRUE),
    prop_includes_gender = mean(includes_gender, na.rm = TRUE),
    prop_includes_pronouns = mean(includes_gendered_pronouns, na.rm = TRUE),
    .groups = "drop"
  )

# Reshape for plotting
inclusion_df <- inclusion_props %>%
  pivot_longer(
    cols = starts_with("prop_"),
    names_to = "inclusion_type",
    names_prefix = "prop_",
    values_to = "proportion"
  ) %>%
  mutate(
    inclusion_type = case_when(
      inclusion_type == "includes_race" ~ "Race",
      inclusion_type == "includes_gender" ~ "Gender",
      inclusion_type == "includes_pronouns" ~ "Gendered Pronouns"
    )
  )

# GPT inclusion plot
inclusion_df %>%
  filter(model == 'GPT') %>%
  filter(!prompt %in% c( "redacted")) %>%
  ggplot(aes(x = prompt, y = proportion, 
             color = prompt, shape = inclusion_type)) +
  geom_point(size = 2.5, alpha = 0.8, position = position_dodge(width = 0.3)) +
  scale_color_manual(values = prompt_colors, labels = prompt_labels) +
  scale_x_discrete(labels = prompt_labels) +
  scale_shape_manual(
    values = c("Race" = 16, "Gender" = 17, "Gendered Pronouns" = 15),
    name = "Inclusion Type"
  ) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  labs(
    x = "",
    y = "p(mentioned | self-call)",
    color = ""
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  guides(color = "none")

ggsave('figures/self-calls/GPT4.1/proportion_included_demographic_bias.png', height = 3, width = 4, dpi = 300)

# Qwen inclusion plot
inclusion_df %>%
  filter(model == 'Qwen') %>%
  filter(!prompt %in% c("redacted")) %>%
  ggplot(aes(x = prompt, y = proportion, 
             color = prompt, shape = inclusion_type)) +
  geom_point(size = 2.5, alpha = 0.8, position = position_dodge(width = 0.3)) +
  scale_color_manual(values = prompt_colors, labels = prompt_labels) +
  scale_x_discrete(labels = prompt_labels) +
  scale_shape_manual(
    values = c("Race" = 16, "Gender" = 17, "Gendered Pronouns" = 15),
    name = "Inclusion Type"
  ) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  labs(
    x = "",
   y = "p(mentioned | self-call)",
   color = ""
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  ) +
  guides(color = "none")

ggsave('figures/self-calls/Qwen2.5-7B/proportion_included_demographic_bias.png', height = 3, width = 4, dpi = 300)
```

Both models almost never self-called when gender and race information were not included in the scenario description (proportion of self-calls for Qwen: `r qwen_counts_df%>%filter(prompt=='removed')%>%pull(prop_tool_use)%>%printnum()`; GPT: `r gpt_counts_df%>%filter(prompt=='removed')%>%pull(prop_tool_use)%>%printnum()`), and rarely when no debiasing instruction was given (Qwen: `r qwen_counts_df%>%filter(prompt=='default')%>%pull(prop_tool_use)%>%printnum()`; GPT: `r gpt_counts_df%>%filter(prompt=='default')%>%pull(prop_tool_use)%>%printnum()`), but self-called reliably more when instructed to ignore race and gender information (Qwen: `r qwen_counts_df%>%filter(prompt=='ignore')%>%pull(prop_tool_use)%>%printnum()`; GPT: `r gpt_counts_df%>%filter(prompt=='ignore')%>%pull(prop_tool_use)%>%printnum()`) or to estimate their decision had they not known the race and gender of the individual (Qwen: `r qwen_counts_df%>%filter(prompt=='if_you_didnt_know')%>%pull(prop_tool_use)%>%printnum()`; GPT: `r gpt_counts_df%>%filter(prompt=='if_you_didnt_know')%>%pull(prop_tool_use)%>%printnum()`). Interestingly, figures for the "Don't discriminate" instruction were low (Qwen: `r qwen_counts_df%>%filter(prompt=='dont_discriminate')%>%pull(prop_tool_use)%>%printnum()`; GPT: `r gpt_counts_df%>%filter(prompt=='dont_discriminate')%>%pull(prop_tool_use)%>%printnum()`).

Focusing on the "Ignore" and "If you didn't know" prompts, GPT's self-calls never included an explicit mention of the individuals' race or gender. Qwen mostly omitted this information, but was less reliable in doing so, both for race (`r inclusion_df%>%filter(model=='Qwen' & prompt=='ignore' & inclusion_type=='Race')%>%pull(proportion)%>%printnum()` and `r inclusion_df%>%filter(model=='Qwen' & prompt=='if_you_didnt_know' & inclusion_type=='Race')%>%pull(proportion)%>%printnum()` out of all tool calls for the "Ignore" and  "If you didn't know" prompts) and for gender (`r inclusion_df%>%filter(model=='Qwen' & prompt=='ignore' & inclusion_type=='Gender')%>%pull(proportion)%>%printnum()` and `r inclusion_df%>%filter(model=='Qwen' & prompt=='if_you_didnt_know' & inclusion_type=='Gender')%>%pull(proportion)%>%printnum()`). GPT mostly omitted gender-neutral pronouns, but failed to do so on `r inclusion_df%>%filter(model=='GPT' & prompt=='ignore' & inclusion_type=='Gendered Pronouns')%>%pull(proportion)%>%printnum()` when instructed to ignore and `r inclusion_df%>%filter(model=='GPT' & prompt=='if_you_didnt_know' & inclusion_type=='Gendered Pronouns')%>%pull(proportion)%>%printnum()` when instructed to estimate its decision if it didn't know the individual's gender and race. These figures were `r inclusion_df%>%filter(model=='Qwen' & prompt=='ignore' & inclusion_type=='Gendered Pronouns')%>%pull(proportion)%>%printnum()` and `r inclusion_df%>%filter(model=='Qwen' & prompt=='if_you_didnt_know' & inclusion_type=='Gendered Pronouns')%>%pull(proportion)%>%printnum()`, respectively, for Qwen. Together, we find that the two models made strategic use of self-calling to simulate a state of ignorance.

