---
title: "anslysis 4 tool use: marginal probabilities"
output: word_document
date: "2025-12-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,      # Don't show code
  warning = FALSE,   # Don't show warnings
  message = FALSE   # Don't show messages

)

library(tidyverse)
library(jsonlite)
library(purrr)
library(stringr)
library(lmerTest)
library(papaja)
library(broom)
library(forcats)
library(patchwork)
library(fs)
library(ggplot2)
library(matrixStats)

```


```{r load data}
source('helpers.R')
source('parse_tool_generations.R')

# Load bias data for both models
Qwen.Tamkin.YN.df <- load_bias_data("qwen")
GPT.Tamkin.YN.df <- load_bias_data("gpt")

# Create response_by_prompt with baseline comparison
Qwen.Tamkin.response_by_prompt <- Qwen.Tamkin.YN.df %>% with_baseline()
GPT.Tamkin.response_by_prompt <- GPT.Tamkin.YN.df %>% with_baseline()

# Calculate mean absolute difference stats
Qwen.Tamkin.mean_abs_diff <- Qwen.Tamkin.response_by_prompt %>% summarize_bias()
GPT.Tamkin.mean_abs_diff <- GPT.Tamkin.response_by_prompt %>% summarize_bias()

# Load tool use data (for parse_tool_generations compatibility)
tool_data <- load_tool_use_data("both")
Qwen.Tamkin.tool_use_combined <- tool_data$qwen
GPT.Tamkin.tool_use_combined <- tool_data$gpt

# Load conditional logit response data (wide format) and pivot to long format
combined_resp_df <- load_tool_response_data("both") %>%
  pivot_conditional_logits()
```


```{r analyze and plot}

combined_yn_df <- Qwen.Tamkin.YN.df %>%
  dplyr::select(model, filled_template,vignette,race,gender,prompt, response)%>%
  rbind(GPT.Tamkin.YN.df %>%
          ungroup()%>%
  dplyr::select(model, filled_template,vignette,race,gender,prompt, response))

combined_full_df <- combined_resp_df %>% 
  filter(
    tool_desc == 'run counterfactual simulation' & 
    as.character(prompt) %in% c('ignore', 'if_you_didnt_know', 'dont_discriminate', 'default')
  ) %>% 
  mutate(after_self_call_response = yes_logit - no_logit) %>%
  select(filled_template, prompt, model_response, after_self_call_response, model) %>%
  merge(
    combined_yn_df)

removed_pyes_df <- combined_yn_df %>%
  filter(prompt=='removed') %>%
  mutate(removed_pyes = exp(response)/(1+exp(response))) %>%
  dplyr::select(model, vignette,race,gender, removed_pyes)

included_pyes_df <- combined_full_df %>%
  mutate(pyes=exp(response)/(1+exp(response)),
         after_self_call_pyes = exp(after_self_call_response)/(1+exp(after_self_call_response))) %>%
  dplyr::select(model, vignette,race,gender,prompt, model_response,after_self_call_pyes)

pyes_df <- removed_pyes_df %>%
  merge(included_pyes_df) %>%
  mutate(product = removed_pyes*after_self_call_pyes) %>%
  group_by(model, prompt,model_response) %>% 
  summarise(weighted_pyes = sum(product)/sum(removed_pyes),
            pyes = mean(after_self_call_pyes))



# ------------------------------------------------------------------------------
# FIGURE 1: Response after simulation, split by internal model response
# ------------------------------------------------------------------------------
# ------------------------------------------------------------------------------
# FIGURE 1: Response after simulation, split by internal model response
# Split into separate figures for each model
# ------------------------------------------------------------------------------

# Qwen figure
combined_full_df %>% 
  filter(model == "Qwen") %>%
  ggplot(aes(x = response, y = after_self_call_response, color = as.factor(model_response))) + 
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  geom_hline(yintercept = 0, linetype = "solid", color = "gray30") +
  geom_vline(xintercept = 0, linetype = "solid", color = "gray30") +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  facet_grid(~ prompt, labeller = labeller(
    prompt = c(
      "ignore" = "Ignore",
      "if_you_didnt_know" = "If You Didn't Know",
      "dont_discriminate" = "Don't Discriminate"
    )
  )) +
  scale_color_manual(
    values = c("No" = "#E41A1C", "Yes" = "#377EB8"),
    labels = c("No" = "Internal model said \"no\"", 
               "Yes" = "Internal model said \"yes\"")
  ) +
  labs(
    x = "Default Response",
    y = "Response Conditioned on Self-Simulation",
    color = NULL,
    title = "Qwen 2.5-7B"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 10, face = "bold"),
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold")
  )

ggsave(
  'figures/self-calls/Qwen2.5-7B/responses_after_simulation_Tamkin.png', 
  height = 4.5, width = 10, dpi = 300
)

# GPT figure
combined_full_df %>% 
  filter(model == "GPT") %>%
  ggplot(aes(x = response, y = after_self_call_response, color = as.factor(model_response))) + 
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  geom_hline(yintercept = 0, linetype = "solid", color = "gray30") +
  geom_vline(xintercept = 0, linetype = "solid", color = "gray30") +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  facet_grid(~ prompt, labeller = labeller(
    prompt = c(
      "ignore" = "Ignore",
      "if_you_didnt_know" = "If You Didn't Know",
      "dont_discriminate" = "Don't Discriminate"
    )
  )) +
  scale_color_manual(
    values = c("No" = "#E41A1C", "Yes" = "#377EB8"),
    labels = c("No" = "Internal model said \"no\"", 
               "Yes" = "Internal model said \"yes\"")
  ) +
  labs(
    x = "Default Response",
    y = "Response Conditioned on Self-Simulation",
    color = NULL,
    title = "GPT-4.1"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 10, face = "bold"),
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold")
  )

ggsave(
  'figures/self-calls/GPT4.1/responses_after_simulation.png', 
  height = 4.5, width = 10, dpi = 300
)

```

Both Qwen and GPT used self-calling reasonably, querying their counterfactual selves when attempting to ignore information or to estimate what they would have decided without it, and mostly removing the to-be-ignored details from the self-call. Still, self-calls were not perfect: the gender of individuals was often revealed by the use of gendered pronouns, and sometimes the model presented the blinded information without including the question itself. A first challenge is therefore to train models to use self-calling more reliably.

Assuming reliable and accurate self-calling, for decisions to be unbiased the model needs to lawfully follow the decision of its blinded counterfactual self. To estimate whether this was the case, we presented the two models with prefilled dialogues in which they were initially presented with a scenario, used self-calling to query what they would have responded to the same scenario with gender and race information removed, and then received a response from their counterfacutal, blinded self: either a "Yes" or a "No". We then measured the relative probability of a "yes" or "no" response following the tool use outcome. 

```{r fit regression model}

regression_model_df <- Qwen.Tamkin.response_by_prompt %>%
  rbind(GPT.Tamkin.response_by_prompt %>%
          dplyr::select(model, vignette,race,gender,prompt, response,removed_response)) %>%
  merge(combined_resp_df %>%
          mutate(after_self_call_response=yes_logit-no_logit) %>%
          filter(tool_desc=='run counterfactual simulation') %>%
          dplyr::select(model, vignette,gender,race,prompt, model_response, after_self_call_response)) %>%
  filter(prompt %in% c('ignore', 'if_you_didnt_know', 'dont_discriminate')) %>%
  mutate(
    model_response = factor(model_response, levels = c("No", "Yes")),
    prompt = factor(prompt, levels = c('dont_discriminate', 'ignore', 'if_you_didnt_know'))
  )

# Fit models for each model × prompt combination
models_by_combo <- regression_model_df %>%
group_by(model, prompt) %>%
  summarise(
    n_obs = n(),
    model_fit = list(lm(after_self_call_response ~ removed_response + race + gender + model_response, data = cur_data())),
    .groups = "drop"
  ) %>%
  mutate(
    tidy_results = map(model_fit, tidy),
    glance_results = map(model_fit, glance)
  )

# Extract coefficients
coefficients_table <- models_by_combo %>%
  select(model, prompt, n_obs, tidy_results) %>%
  unnest(tidy_results) %>%
  mutate(
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01 ~ "**",
      p.value < 0.05 ~ "*",
      TRUE ~ ""
    )
  ) %>%
  select(model, prompt, term, estimate, std.error, statistic, p.value, sig)

print(coefficients_table, n = Inf)

# Extract model fit statistics
model_fit_stats <- models_by_combo %>%
  select(model, prompt, n_obs, glance_results) %>%
  unnest(glance_results) %>%
  select(model, prompt, n_obs, r.squared, adj.r.squared, sigma, AIC, BIC)

print(model_fit_stats)

# Create a summary focusing on race and gender coefficients
race_gender_effects <- coefficients_table %>%
  filter(term %in% c("genderfemale", "raceBlack", "raceHispanic", "raceAsian")) %>%
  arrange(model, prompt, term)

print(race_gender_effects, n = Inf)

# Create a wide format for easier comparison across prompts
coefficients_wide <- coefficients_table %>%
  select(model, prompt, term, estimate, sig) %>%
  mutate(estimate_sig = paste0(sprintf("%.2f", estimate), sig)) %>%
  select(-estimate, -sig) %>%
  pivot_wider(
    names_from = c(model, prompt),
    values_from = estimate_sig,
    names_sep = "_"
  )

print(coefficients_wide, n = Inf)


```







<!-- Responses were mostly positive when the internal model responded with a "Yes" (Qwen: mean response = `r combined_full_df %>% filter(model == "Qwen" & model_response=='Yes')%>%pull(after_self_call_response)%>%mean()%>%printnum()`, SE = `r combined_full_df %>% filter(model == "Qwen" & model_response=='Yes')%>%pull(after_self_call_response)%>%se()%>%printnum()`; `r combined_full_df %>% filter(model == "Qwen" & model_response=='Yes')%>%mutate(positive=after_self_call_response>0)%>%pull(positive)%>%mean()%>%printnum()` of all responses were positive; GPT: mean response = `r combined_full_df %>% filter(model == "GPT" & model_response=='Yes')%>%pull(after_self_call_response)%>%mean()%>%printnum()`, SE = `r combined_full_df %>% filter(model == "GPT" & model_response=='Yes')%>%pull(after_self_call_response)%>%se()%>%printnum()`; `r combined_full_df %>% filter(model == "GPT" & model_response=='Yes')%>%mutate(positive=after_self_call_response>0)%>%pull(positive)%>%mean()%>%printnum()` of all responses were positive), and mostly negative when it responded with a "No" (Qwen: mean response = `r combined_full_df %>% filter(model == "Qwen" & model_response=='No')%>%pull(after_self_call_response)%>%mean()%>%printnum()`, SE = `r combined_full_df %>% filter(model == "Qwen" & model_response=='No')%>%pull(after_self_call_response)%>%se()%>%printnum()`; `r combined_full_df %>% filter(model == "Qwen" & model_response=='No')%>%mutate(negative=after_self_call_response<0)%>%pull(negative)%>%mean()%>%printnum()` of all responses were negative; GPT: mean response = `r combined_full_df %>% filter(model == "GPT" & model_response=='No')%>%pull(after_self_call_response)%>%mean()%>%printnum()`, SE = `r combined_full_df %>% filter(model == "GPT" & model_response=='No')%>%pull(after_self_call_response)%>%se()%>%printnum()`; `r combined_full_df %>% filter(model == "GPT" & model_response=='No')%>%mutate(negative=after_self_call_response<0)%>%pull(negative)%>%mean()%>%printnum()` of all responses were negative).  -->
Qwen almost always deferred to its counterfactual, blinded, self when making decisions. When asked to ignore race and gender information, Qwen responded "Yes" on `r pyes_df%>%filter(model=='Qwen' & prompt=='ignore' & model_response=='Yes')%>%pull(pyes)%>%printnum()` of all trials in which the counterfactual self responded "Yes", and on `r pyes_df%>%filter(model=='Qwen' & prompt=='ignore' & model_response=='No')%>%pull(pyes)%>%printnum()` of the trials in which it responded "No". These figures were `r pyes_df%>%filter(model=='Qwen' & prompt=='dont_discriminate' & model_response=='Yes')%>%pull(pyes)%>%printnum()`/`r pyes_df%>%filter(model=='Qwen' & prompt=='dont_discriminate' & model_response=='No')%>%pull(pyes)%>%printnum()` when instructed not to discriminate, and `r pyes_df%>%filter(model=='Qwen' & prompt=='if_you_didnt_know' & model_response=='Yes')%>%pull(pyes)%>%printnum()`/`r pyes_df%>%filter(model=='Qwen' & prompt=='if_you_didnt_know' & model_response=='No')%>%pull(pyes)%>%printnum()` when estimating what it would have responded had it not known race and gender. Things were different for GPT: whenever its blinded version responded with a "Yes", it almost always deferred to it (probably of responding "Yes" for "ignore": `r pyes_df%>%filter(model=='GPT' & prompt=='ignore' & model_response=='Yes')%>%pull(pyes)%>%printnum()`; "don't discriminate": `r pyes_df%>%filter(model=='GPT' & prompt=='dont_discriminate' & model_response=='Yes')%>%pull(pyes)%>%printnum()`; "if you didn't know": `r pyes_df%>%filter(model=='GPT' & prompt=='if_you_didnt_know' & model_response=='Yes')%>%pull(pyes)%>%printnum()`), but it often overrode the blinded model's decision when it responded with a "No" (probably of responding "Yes" for "ignore": `r pyes_df%>%filter(model=='GPT' & prompt=='ignore' & model_response=='No')%>%pull(pyes)%>%printnum()`; "don't discriminate": `r pyes_df%>%filter(model=='GPT' & prompt=='dont_discriminate' & model_response=='No')%>%pull(pyes)%>%printnum()`; "if you didn't know": `r pyes_df%>%filter(model=='GPT' & prompt=='if_you_didnt_know' & model_response=='No')%>%pull(pyes)%>%printnum()`). Crucially, models overrode their blinded counterparts' decisions when it disagreed with the decision they would have made under true blindness. Generally, this overriding did not happen more often for specific demographics, with the exception that GPT, in the "if you didn't know" intervention, significantly corrected the blinded model's response to be more favourable to Black individuals ($\beta=$ `r race_gender_effects %>% filter(model=='GPT' & prompt=='if_you_didnt_know' & term=='raceBlack')%>%pull(estimate)%>%printnum()`, p `r race_gender_effects %>% filter(model=='GPT' & prompt=='if_you_didnt_know' & term=='raceBlack')%>%pull(p.value)%>%printp(add_equals=T)` from a linear model predicting response from the decision of the blinded model (Yes or No), the response of a blinded model in logit units, race, and gender).

```{r  complete figure}
marginal_response_data <- removed_pyes_df %>%
  merge(included_pyes_df) %>%
  group_by(model, vignette, race, gender, prompt) %>%
  summarise(
    removed_pyes = first(removed_pyes),  # Should be constant within group
    after_self_call_yes = after_self_call_pyes[model_response == 'Yes'][1],
    after_self_call_no = after_self_call_pyes[model_response == 'No'][1],
    marginal_pyes = removed_pyes * after_self_call_yes + (1 - removed_pyes) * after_self_call_no,
    .groups = "drop"
  )  %>%
  mutate(
    marginal_response = log(marginal_pyes)-log(1 - marginal_pyes),
    removed_response = log(removed_pyes)- log(1 - removed_pyes)
  )

create_marginal_figure <- function(original_data, marginal_data, mean_abs_diff_data, 
                                   model_name, output_path) {
  
  # Filter data for this model
  model_original <- original_data %>% filter(model == model_name)
  model_marginal <- marginal_data %>% filter(model == model_name)
  
  # Clean marginal data
  model_marginal_clean <- model_marginal %>%
    filter(
      !is.na(marginal_response) & 
        !is.infinite(marginal_response) &
        !is.na(removed_response) & 
        !is.infinite(removed_response)
    ) %>%
    mutate(
      gender = factor(gender, levels = c("male", "female")),
      race = factor(race, levels = c("white", "Black", "Hispanic", "Asian"))
    )
  
  # Create darker colors (original + 50% black overlay)
  darken_color <- function(color) {
    rgb_vals <- col2rgb(color) / 255
    darkened <- rgb_vals * 0.5  # Equivalent to 50% black overlay
    rgb(darkened[1], darkened[2], darkened[3])
  }
  
  colors <- c(
    dont_discriminate = "#377EB8",
    ignore = "#4DAF4A",
    if_you_didnt_know = "#984EA3"
  )
  
  dark_colors <- sapply(colors, darken_color)
  
# Function to create scatter plot
create_scatter <- function(original_data, marginal_data, prompt_name, prompt_label, color_val) {
  original_subset <- original_data %>% filter(prompt == prompt_name)
  marginal_subset <- marginal_data %>% filter(prompt == prompt_name)
  
  ggplot() +
    geom_hline(yintercept = 0, linetype = "solid", color = "gray30") +
    geom_vline(xintercept = 0, linetype = "solid", color = "gray30") +
    # Original points in background (colored, low opacity)
    geom_point(data = original_subset,
               aes(x = removed_response, y = response), 
               color = color_val, alpha = 0.1, size = 1.5) +
    # Marginal points in foreground (black)
    geom_point(data = marginal_subset,
               aes(x = removed_response, y = marginal_response), 
               color = "black", alpha = 0.8, size = 1.5) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "white") +
    labs(
      title = prompt_label,
      x = "Blind to race and gender",
      y = "Blinded via self-call"
    ) +
    theme_minimal() +
    theme(
      aspect.ratio = 1,
      strip.text = element_text(size = 10, face = "bold"),
      plot.title = element_text(hjust = 0.5, size = 10, face = "bold"),
      plot.margin = margin(3, 1, 1, 1)
    )
}
  
  # Function to extract coefficients
  get_coefs_for_prompt <- function(data, prompt_name) {
    df <- data %>% filter(prompt == prompt_name)
    m4 <- lm(marginal_response ~ removed_response + gender + race, data = df)
    
    tidy(m4) %>%
      filter(term %in% c("genderfemale", "raceBlack", "raceHispanic", "raceAsian")) %>%
      mutate(
        prompt = prompt_name,
        term_clean = case_when(
          term == "genderfemale" ~ "Female",
          term == "raceBlack" ~ "Black",
          term == "raceHispanic" ~ "Hispanic",
          term == "raceAsian" ~ "Asian"
        ),
        term_clean = factor(term_clean, levels = c("Female", "Black", "Hispanic", "Asian")),
        significance = case_when(
          p.value < 0.001 ~ "***",
          p.value < 0.01 ~ "**",
          p.value < 0.05 ~ "*",
          TRUE ~ ""
        ),
        star_y = estimate + std.error + 0.05 * max(abs(estimate))
      )
  }
  
  # Get default coefficients
  get_coefs_for_default <- function(data) {
    df <- data %>% 
      filter(prompt == "default") %>%
      mutate(
        gender = factor(gender, levels = c("male", "female")),
        race = factor(race, levels = c("white", "Black", "Hispanic", "Asian"))
      )
    
    m4 <- lm(response ~ removed_response + gender + race, data = df)
    
    tidy(m4) %>%
      filter(term %in% c("genderfemale", "raceBlack", "raceHispanic", "raceAsian")) %>%
      mutate(
        term_clean = case_when(
          term == "genderfemale" ~ "Female",
          term == "raceBlack" ~ "Black",
          term == "raceHispanic" ~ "Hispanic",
          term == "raceAsian" ~ "Asian"
        ),
        term_clean = factor(term_clean, levels = c("Female", "Black", "Hispanic", "Asian"))
      ) %>%
      select(term_clean, estimate_default = estimate, std.error_default = std.error)
  }
  
  default_coefs <- get_coefs_for_default(model_original)
  
  # Get marginal coefficients
  marginal_coefs <- bind_rows(
    get_coefs_for_prompt(model_marginal_clean, "dont_discriminate"),
    get_coefs_for_prompt(model_marginal_clean, "ignore"),
    get_coefs_for_prompt(model_marginal_clean, "if_you_didnt_know")
  )
  
  # Get coefficients for original interventions (non-self-call)
get_coefs_for_original <- function(data, prompt_name) {
  df <- data %>% 
    filter(prompt == prompt_name) %>%
    mutate(
      gender = factor(gender, levels = c("male", "female")),
      race = factor(race, levels = c("white", "Black", "Hispanic", "Asian"))
    )
  
  m4 <- lm(response ~ removed_response + gender + race, data = df)
  
  tidy(m4) %>%
    filter(term %in% c("genderfemale", "raceBlack", "raceHispanic", "raceAsian")) %>%
    mutate(
      prompt = prompt_name,
      term_clean = case_when(
        term == "genderfemale" ~ "Female",
        term == "raceBlack" ~ "Black",
        term == "raceHispanic" ~ "Hispanic",
        term == "raceAsian" ~ "Asian"
      ),
      term_clean = factor(term_clean, levels = c("Female", "Black", "Hispanic", "Asian"))
    ) %>%
    select(prompt, term_clean, estimate, std.error)
}

# Get original intervention coefficients
original_coefs <- bind_rows(
  get_coefs_for_original(model_original, "dont_discriminate"),
  get_coefs_for_original(model_original, "ignore"),
  get_coefs_for_original(model_original, "if_you_didnt_know")
)

  # Function to create coefficient plot
create_coef_plot <- function(original_coef_data, marginal_coef_data, prompt_name, default_ref, prompt_color) {
  # Get original intervention coefficients
  original_plot_data <- original_coef_data %>% 
    filter(prompt == prompt_name) %>%
    left_join(default_ref, by = "term_clean")
  
  # Get marginal coefficients
  marginal_plot_data <- marginal_coef_data %>% 
    filter(prompt == prompt_name) %>%
    left_join(default_ref, by = "term_clean")
  
  ggplot(marginal_plot_data, aes(x = term_clean)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    # Default in gray background (far left)
    geom_point(aes(y = estimate_default), size = 2, color = "gray50", alpha = 0.3,
               position = position_nudge(x = -0.25)) +
    geom_errorbar(
      aes(ymin = estimate_default - std.error_default, 
          ymax = estimate_default + std.error_default),
      width = 0.2, linewidth = 0.6, color = "gray50", alpha = 0.3,
      position = position_nudge(x = -0.25)
    ) +
    # Original intervention in color with low opacity (middle left)
    geom_point(data = original_plot_data, aes(y = estimate), 
               size = 2, color = prompt_color, alpha = 0.3,
               position = position_nudge(x = -0.1)) +
    geom_errorbar(
      data = original_plot_data,
      aes(ymin = estimate - std.error, ymax = estimate + std.error),
      width = 0.2, linewidth = 0.6, color = prompt_color, alpha = 0.3,
      position = position_nudge(x = -0.1)
    ) +
    # Marginal (self-call) in black (right)
    geom_point(aes(y = estimate), size = 2, color = "black",
               position = position_nudge(x = 0.1)) +
    geom_errorbar(
      aes(ymin = estimate - std.error, 
          ymax = estimate + std.error),
      width = 0.2, linewidth = 0.6, color = "black",
      position = position_nudge(x = 0.1)
    ) +
    geom_text(
      aes(label = significance, y = star_y),
      size = 4, color = "black", fontface = "bold", vjust = 0,
      position = position_nudge(x = 0.1)
    ) +
    labs(x = NULL, y = "β") +
    theme_minimal() +
    coord_cartesian(clip = "off") +
    theme(
      aspect.ratio = 0.3,
      axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
      axis.title.y = element_text(size = 8),
      panel.grid.major.x = element_blank(),
      plot.margin = margin(t = 15, r = 1, b = 2, l = 1)
    )
}

  # Create scatter plots with darker colors
# Create scatter plots with original in background
# Create scatter plots with original colors at alpha=0.1
s1 <- create_scatter(model_original, model_marginal_clean, "dont_discriminate", 
                     "Don't Discriminate", "#377EB8")
s2 <- create_scatter(model_original, model_marginal_clean, "ignore", 
                     "Ignore", "#4DAF4A")
s3 <- create_scatter(model_original, model_marginal_clean, "if_you_didnt_know", 
                     "If You Didn't Know", "#984EA3")
  
  # Create coefficient plots with darker colors
# Create coefficient plots - pass the full dataframes, function will filter
c1 <- create_coef_plot(original_coefs, marginal_coefs, "dont_discriminate", default_coefs, "#377EB8")
c2 <- create_coef_plot(original_coefs, marginal_coefs, "ignore", default_coefs, "#4DAF4A")
c3 <- create_coef_plot(original_coefs, marginal_coefs, "if_you_didnt_know", default_coefs, "#984EA3")
  
  # Combine scatter + coef
  col1 <- (s1 / c1) + plot_layout(heights = c(3, 1))
  col2 <- (s2 / c2) + plot_layout(heights = c(3, 1))
  col3 <- (s3 / c3) + plot_layout(heights = c(3, 1))
  
  # Calculate mean abs diff for marginal
  marginal_mean_abs_diff <- model_marginal_clean %>%
    group_by(prompt) %>%
    summarise(
      mean_abs_diff = mean(abs(marginal_response - removed_response)),
      se_abs_diff = sd(abs(marginal_response - removed_response)) / sqrt(n()),
      type = "marginal",
      .groups = "drop"
    )
  
  # Get original mean abs diff
  original_mean_abs_diff <- mean_abs_diff_data %>%
    mutate(type = "original")
  
  # Combine
  combined_mean_abs_diff <- bind_rows(original_mean_abs_diff, marginal_mean_abs_diff) %>%
    filter(prompt %in% c('default', 'dont_discriminate', 'ignore', 'if_you_didnt_know'))
  
  # Separate for plotting
  original_data_bar <- combined_mean_abs_diff %>% filter(type == "original")
  marginal_data_bar <- combined_mean_abs_diff %>% filter(type == "marginal")
  
 # Create barplot with new order
p_bar <- original_data_bar %>%
  mutate(
    fill_color = case_when(
      prompt == "default" ~ "gray",
      prompt == "dont_discriminate" ~ "#377EB8",
      prompt == "ignore" ~ "#4DAF4A",
      prompt == "if_you_didnt_know" ~ "#984EA3"
    ),
    prompt = factor(prompt, levels = c("if_you_didnt_know","dont_discriminate",  "ignore", "default")),
    label_text = case_when(
      prompt == "default" ~ "Default",
      prompt == "dont_discriminate" ~ "Don't Discriminate",
      prompt == "ignore" ~ "Ignore",
      prompt == "if_you_didnt_know" ~ "If You Didn't Know"
    )
  ) %>%
  ggplot(aes(x = prompt, y = mean_abs_diff)) +
  # Original bars
  geom_bar(aes(fill = fill_color), stat = "identity", width = 0.7) +
  scale_fill_identity() +
  # Marginal bars overlaid in black
  geom_bar(data = marginal_data_bar %>% 
             mutate(prompt = factor(prompt, levels = c("if_you_didnt_know","dont_discriminate",  "ignore", "default"))),
           stat = "identity", width = 0.7, 
           fill = "black") +
  # Error bars
  geom_errorbar(aes(ymin = mean_abs_diff - se_abs_diff, 
                    ymax = mean_abs_diff + se_abs_diff),
                width = 0.2, linewidth = 0.8) +
  geom_errorbar(data = marginal_data_bar %>% 
                  mutate(prompt = factor(prompt, levels = c("if_you_didnt_know","dont_discriminate",  "ignore", "default"))),
                aes(ymin = mean_abs_diff - se_abs_diff, 
                    ymax = mean_abs_diff + se_abs_diff),
                width = 0.2, linewidth = 0.8) +
  # Text labels
  geom_text(
    aes(y = 0.3, label = label_text),
    hjust = 0, size = 3, fontface = "bold", color = "white"
  ) +
  scale_y_continuous(expand = c(0, 0), 
                     limits = c(0, max(combined_mean_abs_diff$mean_abs_diff) * 1.1)) +
  labs(x = NULL, y = "mean absolute difference from true blindness") +
  coord_flip() +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank()
  )  
  # Combine everything
  final_plot <- (col1 | col2) / (col3 | p_bar)
  
  # Save
  ggsave(output_path, plot = final_plot, height = 7, width = 6, dpi = 300)
  
  return(list(marginal_coefs = marginal_coefs, default_coefs = default_coefs))
}

# Apply to GPT
gpt_results <- create_marginal_figure(
  GPT.Tamkin.response_by_prompt,
  marginal_response_data,
  GPT.Tamkin.mean_abs_diff,
  "GPT",
  'figures/tamkin/GPT4.1/marginal_complete_figure.png'
)

# Apply to Qwen
qwen_results <- create_marginal_figure(
  Qwen.Tamkin.response_by_prompt,
  marginal_response_data,
  Qwen.Tamkin.mean_abs_diff,
  "Qwen",
  'figures/tamkin/Qwen2.5-7B/marginal_complete_figure.png'
)

```
 
Despite these imperfections, access to the response of a counterfactual, blinded, self, produced responses that were more aligned with the resposnes of a truly blinded model than any of the other tested interventions. Indeed, the mean absolute difference relative to a truly blinded response was lower than default in all three interventions (black bars in Fig. X), and responses were no longer significantly biased in favour, or against, any specific demographic.


```{r default}

create_default_selfcall_comparison <- function(original_data, marginal_data, output_path) {
  
  # Colors for each model
  gpt_color <- "darkgray"  
  qwen_color <- "darkgray"  
  
  # Function to create scatter plot for one model
  create_scatter <- function(original_data, marginal_data, model_name, model_label, color_val) {
    original_subset <- original_data %>% 
      filter(model == model_name, prompt == "default")
    
    marginal_subset <- marginal_data %>% 
      filter(model == model_name, prompt == "default") %>%
      filter(
        !is.na(marginal_response) & 
        !is.infinite(marginal_response) &
        !is.na(removed_response) & 
        !is.infinite(removed_response)
      )
    
    ggplot() +
      geom_hline(yintercept = 0, linetype = "solid", color = "gray30") +
      geom_vline(xintercept = 0, linetype = "solid", color = "gray30") +
      # Original points in background (colored, alpha = 0.1)
      geom_point(data = original_subset,
                 aes(x = removed_response, y = response), 
                 color = color_val, alpha = 0.7, size = 1.5) +
      # Marginal points in foreground (black)
      geom_point(data = marginal_subset,
                 aes(x = removed_response, y = marginal_response), 
                 color = "black", alpha = 0.7, size = 1.5) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "white") +
      labs(
        title = model_label,
        x = "Blind to race and gender",
        y = "Blinded via self-call"
      ) +
      theme_minimal() +
      theme(
        aspect.ratio = 1,
        plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
        plot.margin = margin(3, 1, 1, 1)
      )
  }
  
  # Function to extract coefficients for original
  get_coefs_original <- function(data, model_name) {
    df <- data %>% 
      filter(model == model_name, prompt == "default") %>%
      mutate(
        gender = factor(gender, levels = c("male", "female")),
        race = factor(race, levels = c("white", "Black", "Hispanic", "Asian"))
      )
    
    m4 <- lm(response ~ removed_response + gender + race, data = df)
    
    tidy(m4) %>%
      filter(term %in% c("genderfemale", "raceBlack", "raceHispanic", "raceAsian")) %>%
      mutate(
        model = model_name,
        term_clean = case_when(
          term == "genderfemale" ~ "Female",
          term == "raceBlack" ~ "Black",
          term == "raceHispanic" ~ "Hispanic",
          term == "raceAsian" ~ "Asian"
        ),
        term_clean = factor(term_clean, levels = c("Female", "Black", "Hispanic", "Asian"))
      ) %>%
      select(model, term_clean, estimate, std.error)
  }
  
  # Function to extract coefficients for marginal
  get_coefs_marginal <- function(data, model_name) {
    df <- data %>% 
      filter(model == model_name, prompt == "default") %>%
      filter(
        !is.na(marginal_response) & 
        !is.infinite(marginal_response) &
        !is.na(removed_response) & 
        !is.infinite(removed_response)
      ) %>%
      mutate(
        gender = factor(gender, levels = c("male", "female")),
        race = factor(race, levels = c("white", "Black", "Hispanic", "Asian"))
      )
    
    m4 <- lm(marginal_response ~ removed_response + gender + race, data = df)
    
    tidy(m4) %>%
      filter(term %in% c("genderfemale", "raceBlack", "raceHispanic", "raceAsian")) %>%
      mutate(
        model = model_name,
        term_clean = case_when(
          term == "genderfemale" ~ "Female",
          term == "raceBlack" ~ "Black",
          term == "raceHispanic" ~ "Hispanic",
          term == "raceAsian" ~ "Asian"
        ),
        term_clean = factor(term_clean, levels = c("Female", "Black", "Hispanic", "Asian")),
        significance = case_when(
          p.value < 0.001 ~ "***",
          p.value < 0.01 ~ "**",
          p.value < 0.05 ~ "*",
          TRUE ~ ""
        ),
        star_y = estimate + std.error + 0.05 * max(abs(estimate))
      )
  }
  
  # Get coefficients for both models
  gpt_coefs_original <- get_coefs_original(original_data, "GPT")
  qwen_coefs_original <- get_coefs_original(original_data, "Qwen")
  
  gpt_coefs_marginal <- get_coefs_marginal(marginal_data, "GPT")
  qwen_coefs_marginal <- get_coefs_marginal(marginal_data, "Qwen")
  
  # Function to create coefficient plot
  create_coef_plot <- function(original_coefs, marginal_coefs, model_name, color_val) {
    original_plot_data <- original_coefs %>% filter(model == model_name)
    marginal_plot_data <- marginal_coefs %>% filter(model == model_name)
    
    ggplot(marginal_plot_data, aes(x = term_clean)) +
      geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
      # Original in color with low opacity (left)
      geom_point(data = original_plot_data, aes(y = estimate), 
                 size = 2, color = color_val, alpha = 0.8,
                 position = position_nudge(x = -0.15)) +
      geom_errorbar(
        data = original_plot_data,
        aes(ymin = estimate - std.error, ymax = estimate + std.error),
        width = 0.2, linewidth = 0.6, color = color_val, alpha = 0.3,
        position = position_nudge(x = -0.15)
      ) +
      # Marginal in black (right)
      geom_point(aes(y = estimate), size = 2, color = "black",
                 position = position_nudge(x = 0.15)) +
      geom_errorbar(
        aes(ymin = estimate - std.error, 
            ymax = estimate + std.error),
        width = 0.2, linewidth = 0.6, color = "black",
        position = position_nudge(x = 0.15)
      ) +
      geom_text(
        aes(label = significance, y = star_y),
        size = 4, color = "black", fontface = "bold", vjust = 0,
        position = position_nudge(x = 0.15)
      ) +
      labs(x = NULL, y = "β") +
      theme_minimal() +
      coord_cartesian(clip = "off") +
      theme(
        aspect.ratio = 0.3,
        axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
        axis.title.y = element_text(size = 8),
        panel.grid.major.x = element_blank(),
        plot.margin = margin(t = 15, r = 1, b = 2, l = 1)
      )
  }
  
  # Create plots for both models
  s_gpt <- create_scatter(original_data, marginal_data, "GPT", "GPT-4.1", gpt_color)
  s_qwen <- create_scatter(original_data, marginal_data, "Qwen", "Qwen2.5-7B-Instruct", qwen_color)
  
  all_coefs_original <- bind_rows(gpt_coefs_original, qwen_coefs_original)
  all_coefs_marginal <- bind_rows(gpt_coefs_marginal, qwen_coefs_marginal)
  
  c_gpt <- create_coef_plot(all_coefs_original, all_coefs_marginal, "GPT", gpt_color)
  c_qwen <- create_coef_plot(all_coefs_original, all_coefs_marginal, "Qwen", qwen_color)
  
  # Combine scatter + coef for each model
  col_gpt <- (s_gpt / c_gpt) + plot_layout(heights = c(3, 1))
  col_qwen <- (s_qwen / c_qwen) + plot_layout(heights = c(3, 1))
  
  # Combine side by side
  final_plot <- col_qwen | col_gpt 
  
  # Save
  ggsave(output_path, plot = final_plot, height = 5, width = 10, dpi = 300)
  
  return(list(
    gpt_coefs_original = gpt_coefs_original,
    qwen_coefs_original = qwen_coefs_original,
    gpt_coefs_marginal = gpt_coefs_marginal,
    qwen_coefs_marginal = qwen_coefs_marginal
  ))
}

# Create the comparison figure
default_comparison_results <- create_default_selfcall_comparison(
  bind_rows(GPT.Tamkin.response_by_prompt, Qwen.Tamkin.response_by_prompt),
  marginal_response_data,
  'figures/tamkin/default_selfcall_comparison.png'
)

```